---
title: "Exploratory Analysis"
output: html_document
---

## Project title

Give your project a concise, interesting title that summarizes the entirety of your project

### Authors

| Name     | Email      |
|:-------:|:--------:|
| Bonnie   | bonche22@uw.edu |
| Ling Lin | lingl3@uw.edu.  |
| Yuchen AN| ayc0816@uw.edu. |
| Mandy    | rwu1@uw.edu     |

### Date

Winter 2022

### Abstract

No more than three sentences that summarize your project. Focus on the very most important aspects. For example: (1) "Our main question is .... This question is important because .... To address the question, we will ...." (2) "We are concerned with ..., because .... To address this concern, we plan to ...." (3) "Consider that .... This is important because .... Accordingly, we plan to ...."

### Keywords

3-5 keywords that summarize your project.
(e.g., "Keywords: human physiology; bicycle exercise; Gen Z; times-series data")

### Introduction  

Briefly introduce your project.  Include 3-5 research questions. What motivates the questions? Why are they important? (at least 200 words)

### Related Work  

Describe your topic and related work in this space. You must include 3 citations to related work (URLs to similar work, high quality articles from the popular press, research papers, etc. ) Please use a standard citation style of your choice. (at least 200 words)

### The Dataset

Where did you find the data? Please include a link to the data source  
Who collected the data?  
How was the data collected or generated?  
Why was the data collected?  
How many observations (rows) are in your data?  
How many features (columns) are in the data?  
What, if any, ethical questions or questions of power do you need to consider when working with this data?  
What are possible limitations or problems with this data?   (at least 200 words)

### Implications

Assuming you answer your research questions, briefly describe the expected or possible implications for technologists, designers, and policymakers. (at least 150 words)

### Limitations & Challenges
What challenges or limitations might you need to address with your project idea more broadly? Briefly discuss. (at least 150 words)

### Summary Information

Write a summary paragraph of findings that includes the 5 values calculated from your summary information R script

### Table

Include a table of aggregate information

Describe why you included the table and what information it reveals

### Chart 1

Include a chart

Describe why you chose this chart and what information it reveals

Here's an example of how to run an R script inside an RMarkdown file:

```{r, echo=FALSE, code = readLines("chart1_example.R")}
print("hello world")
```

### Chart 2

Include a chart
```{r, echo = FALSE, warning = FALSE, message = FALSE, code = readLines("chart 2.R")}
```

Describe why you chose this chart and what information it reveals

### Chart 3

I chose to create this chart to analyze the correlation between a college's acceptance rate and 4-year graduation rate to analyze whether the selectivity of an institution has an impact on the success of their students to graduate on time. I also added a differentiation for private versus public institutions to show whether more selective colleges are generally public or private. 

After creating the chart, it reveals that there is a slight correlation between the two factors. The most selective colleges are almost all private institutions, and they tend to have a higher graduation rate. However, there are also a handful of data points that show selective institutions that have a lower 4-year graduation rate. Lastly, the chart shows that most colleges tend to fall between a 50-80% acceptance rate with a 4-year graduation rate between 15-50%. 

```{r, echo = FALSE, warning=FALSE, message=FALSE}
#Load libraries 
library("ggplot2")
library("plotly")
library("dplyr")

college_df <- read.csv("https://raw.githubusercontent.com/info-201a-wi23/exploratory-analysis-mandyrwu/main/Data-Table%201.csv", stringsAsFactors = FALSE)

admissions_data <- college_df %>%
  select(Name, Applicants.total, Admissions.total, Graduation.rate...Bachelor.degree.within.4.years..total, Control.of.institution)

admissions_data <- mutate(admissions_data, "Acceptance.rate" = (Admissions.total/Applicants.total) * 100)

ggplot(data = admissions_data) +
  geom_point(mapping = aes(x = Acceptance.rate, y = Graduation.rate...Bachelor.degree.within.4.years..total, color = Control.of.institution)) +
  labs(title = "Colleges in the US: Accaptance Rate vs Undergrad Graduation Rate",
       x = "Acceptance Rate (%)",
       y = "Graduation Rate with Bachelor Degreen within 4 Years (%)")
```

Include a chart

Describe why you chose this chart and what information it reveals
